---
title: "Mixed-Effects Modelling for Ergonomics Research"
subtitle: "A beginner-oriented walkthrough of the worked examples"

author:
  - name: Michael W. B. Watterworth
    degrees: MHSc
    orcid: 0000-0001-5227-4554
    url: "https://watterworth.github.io"
    corresponding: true
    email: "michael.watterworth@ontariotechu.net"
    affiliations:
      - name: Ontario Tech University
        department: Faculty of Health Sciences
        city: Oshawa
        state: Ontario
        country: Canada

  - name: Nicholas J. La Delfa
    degrees: PhD
    orcid: 0000-0002-6928-5441
    url: "https://healthsciences.ontariotechu.ca/people/faculty/nick-la-delfa.php"
    corresponding: true
    email: "nicholas.ladelfa@ontariotechu.ca"
    affiliations:
      - name: Ontario Tech University
        department: Faculty of Health Sciences
        city: Oshawa
        state: Ontario
        country: Canada

  - name: Michael A. Rotondi
    degrees: PhD
    url: "https://discover.academics.yorku.ca/Michael.Rotondi"
    corresponding: true
    email: "mrotondi@yorku.ca"
    affiliations:
      - name: York University
        department: School of Kinesiology and Health Science
        city: Toronto
        state: Ontario
        country: Canada

output-file: index.html

format:
  html:
    toc: true
    code-copy: true

execute:
  warning: false
  message: false
---

## How to use this tutorial

This tutorial accompanies the worked examples in *Moving Beyond Repeated Measures ANOVA: A Practical Guide to Mixed-Effects Modeling for Ergonomics Research*. It assumes only basic familiarity with R. Each object is created before it is used, and the code can be run from top to bottom.

Open `lmm-ergonomics-guide.Rproj` in RStudio before running the code. This sets the project folder as the working directory, which allows R to find the supplied dataset.

```{r}
#| label: setup
#| include: false

options(width = 120)
```

## 1. Load the packages

Packages extend what R can do. The `library()` function makes an installed package available in the current R session.

```{r}
#| label: load-packages

library(lme4)        # Fits linear mixed-effects models
library(lmerTest)    # Adds F-tests and denominator degrees of freedom
library(emmeans)     # Estimates model-based means, trends, and contrasts
library(ez)          # Fits the comparison repeated-measures ANOVA
library(performance) # Provides model-diagnostic figures
library(ggplot2)     # Creates figures
library(patchwork)   # Combines multiple figure panels
library(sjPlot)      # Provides various model plots
```

If R reports that a package is missing, install it once with `install.packages("package_name")`, and then run the corresponding `library()` line again.

## 2. Preparing and exploring the data

The illustrative dataset was simulated using the optional code below. The code is included for transparency but is not executed when the tutorial is rendered. Readers can expand the dropdown to see exactly how the observations were generated.

```{r}
#| label: dataset-generation-code
#| eval: false
#| code-fold: true
#| code-summary: "Show dataset generation code"

set.seed(1001)

n_participants <- 14
bins <- c("Near", "Mid", "Far")

sex_by_participant <- rep(
  c("Male", "Female"),
  length.out = n_participants
)

Participant <- factor(
  rep(
    paste0("P", sprintf("%02d", 1:n_participants)),
    each = length(bins)
  )
)

Sex <- factor(
  rep(sex_by_participant, each = length(bins)),
  levels = c("Male", "Female")
)

Reach_bin <- factor(
  rep(bins, times = n_participants),
  levels = bins
)

sample_in_bin <- function(bin) {
  switch(
    bin,
    "Near" = sample(20:35, 1),
    "Mid" = sample(36:55, 1),
    "Far" = sample(56:70, 1)
  )
}

Reach_cm <- vapply(
  as.character(Reach_bin),
  sample_in_bin,
  numeric(1)
)
Reach_cm <- as.integer(Reach_cm)

re_participant <- rnorm(
  n_participants,
  mean = 0,
  sd = 1
)

re_lookup <- setNames(
  re_participant,
  levels(Participant)
)

rand_int <- re_lookup[as.character(Participant)]

beta0 <- 9
b_reach_male <- 0.12
b_reach_female <- b_reach_male - 0.05
b_female_int <- -1.0
sigma_eps <- 1

slope_by_sex <- ifelse(
  Sex == "Male",
  b_reach_male,
  b_reach_female
)

int_by_sex <- ifelse(
  Sex == "Male",
  beta0,
  beta0 + b_female_int
)

Shoulder_Moment_Nm <- int_by_sex +
  slope_by_sex * Reach_cm +
  rand_int +
  rnorm(
    length(Reach_cm),
    mean = 0,
    sd = sigma_eps
  )

Shoulder_Moment_Nm <- round(
  pmax(Shoulder_Moment_Nm, 0),
  2
)

ergo_simple <- data.frame(
  Participant,
  Sex,
  Reach_cm,
  Shoulder_Moment_Nm,
  Reach_bin
)

reach_center_cm <- mean(
  ergo_simple$Reach_cm,
  na.rm = TRUE
)

ergo_simple$Reach_c <-
  ergo_simple$Reach_cm - reach_center_cm
ergo_simple$Reach_c10 <- ergo_simple$Reach_c / 10
ergo_simple$Reach_10 <- ergo_simple$Reach_cm / 10

write.csv(
  ergo_simple,
  "data/ergo_simple_shoulder_moment_balanced.csv",
  row.names = FALSE
)
```

The resulting CSV file is already included with the package. `read.csv()` imports it as a data frame called `ergo_simple`.

```{r}
#| label: import-data

ergo_simple <- read.csv(
  "data/ergo_simple_shoulder_moment_balanced.csv"
)

head(ergo_simple)
```

Participant, sex, and reach bin are categories rather than continuous numbers. Converting them to factors tells R to treat them as categorical variables. The order supplied to `levels` also establishes the reference level and display order.

```{r}
#| label: prepare-factors

ergo_simple$Participant <- factor(ergo_simple$Participant)

ergo_simple$Sex <- factor(
  ergo_simple$Sex,
  levels = c("Male", "Female")
)

ergo_simple$Reach_bin <- factor(
  ergo_simple$Reach_bin,
  levels = c("Near", "Mid", "Far")
)
```

The following figure displays the same observations in the two forms used in Example 1. Panel A preserves the measured reach distance as a continuous variable. Panel B categorizes the observations into the prescribed Near, Mid, and Far reach conditions.

```{r}
#| label: fig-continuous-versus-factor
#| fig-cap: "Continuous and factor-based representations of reach distance and shoulder moment. Both panels display the same observations. Panel A preserves reach distance as a continuous variable, whereas Panel B categorizes reach into Near, Mid, and Far conditions."
#| fig-width: 12
#| fig-height: 5.8
#| out-width: 100%
#| fig-align: center
#| code-fold: true
#| code-summary: "Show figure-generation code"

figure_theme <- theme_classic(base_size = 16) +
  theme(
    strip.background = element_rect(
      fill = "grey92",
      colour = "black",
      linewidth = 0.7
    ),
    strip.text = element_text(
      face = "bold",
      size = 18,
      margin = margin(t = 8, b = 8)
    ),
    axis.title = element_text(
      face = "bold",
      size = 19
    ),
    axis.text = element_text(
      size = 16,
      colour = "black"
    ),
    legend.position = "bottom",
    legend.title = element_text(
      face = "bold",
      size = 17
    ),
    legend.text = element_text(size = 16),
    plot.margin = margin(t = 8, r = 12, b = 8, l = 12)
  )

continuous_data <- ergo_simple
continuous_data$Panel <- "A. Continuous Analysis"

continuous_plot <- ggplot(
  continuous_data,
  aes(
    x = Reach_cm,
    y = Shoulder_Moment_Nm,
    colour = Sex
  )
) +
  geom_point(size = 3.4, alpha = 0.75) +
  facet_wrap(~ Panel) +
  scale_colour_viridis_d(option = "D", end = 0.85) +
  guides(
    colour = guide_legend(nrow = 1, byrow = TRUE)
  ) +
  labs(
    x = "Reach distance (cm)",
    y = "Shoulder moment (Nm)",
    colour = "Sex"
  ) +
  figure_theme

factor_data <- ergo_simple
factor_data$Panel <- "B. Factor-Based Analysis"

factor_plot <- ggplot(
  factor_data,
  aes(
    x = Reach_bin,
    y = Shoulder_Moment_Nm
  )
) +
  geom_boxplot(
    aes(fill = Sex, colour = Sex),
    width = 0.65,
    alpha = 0.55,
    linewidth = 1,
    outlier.shape = NA,
    position = position_dodge(width = 0.75)
  ) +
  geom_point(
    aes(colour = Sex),
    size = 2.8,
    alpha = 0.75,
    position = position_jitterdodge(
      jitter.width = 0.10,
      jitter.height = 0,
      dodge.width = 0.75
    )
  ) +
  facet_wrap(~ Panel) +
  scale_colour_viridis_d(option = "D", end = 0.85) +
  scale_fill_viridis_d(option = "D", end = 0.85) +
  guides(
    fill = "none",
    colour = guide_legend(nrow = 1, byrow = TRUE)
  ) +
  labs(
    x = "Reach condition",
    y = "Shoulder moment (Nm)",
    colour = "Sex"
  ) +
  figure_theme

analysis_comparison_figure <-
  continuous_plot +
  factor_plot +
  plot_layout(
    guides = "collect",
    widths = c(1, 1)
  ) &
  theme(legend.position = "bottom")

analysis_comparison_figure
```

*Note.* Points represent individual observations, and colours distinguish male and female participants. In Panel B, the horizontal line within each box represents the median, the box represents the interquartile range, and the whiskers extend to the most extreme values within 1.5 times the interquartile range.

## 3. Example 1: Factor-based rANOVA versus continuous LMMs

The first analysis treats the prescribed reach conditions as a categorical within-participant factor. The second replaces those categories with the corresponding continuous reach measurements.

### 3.1 Factor-based LMM and rANOVA

### Check the repeated-measures structure

The table below should contain a 1 in every cell. This confirms that each participant has one observation in every reach bin.

```{r}
#| label: check-balanced-design

with(
  ergo_simple,
  table(Participant, Reach_bin)
)
```

### Fit the factor-based LMM

The outcome appears to the left of `~`. The fixed effects appear to its right. `Reach_bin * Sex` includes the two main effects and their interaction. `(1 | Participant)` adds a random intercept, allowing participants to have different baseline shoulder moments.

```{r}
#| label: fit-factor-lmm

model_bins <- lmer(
  Shoulder_Moment_Nm ~ Reach_bin * Sex +
    (1 | Participant),
  data = ergo_simple,
  REML = TRUE
)
```

The next command provides the Type III F-tests used in the article. `ddf = "Satterthwaite"` requests Satterthwaite-approximated denominator degrees of freedom.

```{r}
#| label: test-factor-lmm

anova_bins_lmm <- anova(
  model_bins,
  type = 3,
  ddf = "Satterthwaite"
)

print(anova_bins_lmm)
```

The full coefficient table can be obtained with `summary(model_bins)`. It is not printed in this walkthrough because the reference-coded coefficients are not used for the primary interpretation. The command remains in the complete R script.

### Fit the equivalent rANOVA

This analysis uses the same complete, balanced data. `dv` identifies the outcome, `wid` identifies the participant, `within` identifies the repeated factor, and `between` identifies the between-participant factor.

```{r}
#| label: fit-repeated-measures-anova

anova_bins <- ezANOVA(
  data = ergo_simple,
  dv = Shoulder_Moment_Nm,
  wid = Participant,
  within = Reach_bin,
  between = Sex,
  detailed = TRUE,
  type = 3
)

print(anova_bins$ANOVA)
```

Under these ideal conditions and equivalent model specifications, the rANOVA and factor-based LMM produce the same tests for the fixed effects.

### Estimate and compare the model-based means

`emmeans()` estimates shoulder moment for each reach bin within each sex. These are estimated marginal means, or EMMs.

```{r}
#| label: estimate-reach-bin-emms

emm_bins <- emmeans(
  model_bins,
  ~ Reach_bin | Sex,
  lmer.df = "satterthwaite"
)

print(emm_bins)
```

`pairs()` compares the three reach bins within each sex. The Tukey method adjusts for the three pairwise comparisons in each family.

```{r}
#| label: compare-reach-bin-emms

pairs_bins <- pairs(
  emm_bins,
  adjust = "tukey"
)

print(pairs_bins)
```

The EMMs can also be plotted so the figure displays the same model-based quantities used in the comparisons.

```{r}
#| label: plot-reach-bin-emms

plot(emm_bins, comparisons = TRUE)
```

The default plot is convenient for quickly examining the EMMs. For greater control over the figure's appearance, both the EMMs and their contrasts can be converted to ordinary data frames using `as.data.frame()` and plotted with `ggplot2`. The example below displays the EMMs and their 95% confidence intervals in separate panels for males and females. The significance brackets are generated from the same Tukey-adjusted pairwise comparisons reported above.

```{r}
#| label: plot-reach-bin-emms-with-ggplot
#| fig-cap: "Estimated marginal means and 95% confidence intervals for shoulder moment across reach conditions within each sex. Asterisks indicate Tukey-adjusted pairwise comparisons with p < .05."
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 100%
#| fig-align: center
#| code-fold: true
#| code-summary: "Show customized EMM plotting code"

# Convert the emmeans results to ordinary data frames.
emm_bins_df <- as.data.frame(emm_bins)
pairs_bins_df <- as.data.frame(pairs_bins)

# Assign numeric positions to the reach conditions.
reach_order <- c("Near", "Mid", "Far")

emm_bins_df$x_position <- match(
  as.character(emm_bins_df$Reach_bin),
  reach_order
)

# Retain comparisons with a Tukey-adjusted p-value below .05.
significant_pairs <- pairs_bins_df[
  pairs_bins_df$p.value < 0.05,
]

# Separate each contrast into its two reach conditions.
contrast_parts <- strsplit(
  as.character(significant_pairs$contrast),
  " - ",
  fixed = TRUE
)

significant_pairs$x_start <- vapply(
  contrast_parts,
  function(x) match(x[1], reach_order),
  integer(1)
)

significant_pairs$x_end <- vapply(
  contrast_parts,
  function(x) match(x[2], reach_order),
  integer(1)
)

# Use one significance indicator for all comparisons with p < .05.
significant_pairs$significance <- "*"

# Determine the vertical position of each bracket.
significant_pairs$bracket_y <- NA_real_

for (current_sex in unique(as.character(significant_pairs$Sex))) {

  row_indices <- which(
    as.character(significant_pairs$Sex) == current_sex
  )

  comparison_span <-
    significant_pairs$x_end[row_indices] -
    significant_pairs$x_start[row_indices]

  # Place the widest bracket lowest and shorter brackets above it.
  row_indices <- row_indices[
    order(comparison_span, decreasing = TRUE)
  ]

  panel_maximum <- max(
    emm_bins_df$upper.CL[
      as.character(emm_bins_df$Sex) == current_sex
    ]
  )

  significant_pairs$bracket_y[row_indices] <-
    panel_maximum + seq_along(row_indices) * 0.8
}

# Plot the EMMs, confidence intervals, and adjusted comparisons.
emm_bins_figure <- ggplot(
  emm_bins_df,
  aes(
    x = x_position,
    y = emmean,
    group = Sex,
    colour = Sex
  )
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.08,
    linewidth = 0.8
  ) +
  geom_segment(
    data = significant_pairs,
    aes(
      x = x_start,
      xend = x_end,
      y = bracket_y,
      yend = bracket_y
    ),
    inherit.aes = FALSE,
    colour = "black"
  ) +
  geom_segment(
    data = significant_pairs,
    aes(
      x = x_start,
      xend = x_start,
      y = bracket_y,
      yend = bracket_y - 0.15
    ),
    inherit.aes = FALSE,
    colour = "black"
  ) +
  geom_segment(
    data = significant_pairs,
    aes(
      x = x_end,
      xend = x_end,
      y = bracket_y,
      yend = bracket_y - 0.15
    ),
    inherit.aes = FALSE,
    colour = "black"
  ) +
  geom_text(
    data = significant_pairs,
    aes(
      x = (x_start + x_end) / 2,
      y = bracket_y + 0.12,
      label = significance
    ),
    inherit.aes = FALSE,
    colour = "black",
    size = 5
  ) +
  facet_wrap(~ Sex) +
  scale_colour_viridis_d(
    option = "D",
    end = 0.85
  ) +
  labs(
    x = "Reach Condition",
    y = "Estimated Shoulder Moment (Nm)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )

print(emm_bins_figure)
```

Once converted to a data frame, the EMMs can be plotted using `ggplot2` or exported for visualization in another plotting program. Regardless of the software used, the plotted estimates, confidence intervals, and significance indicators should be based on the model-derived EMMs and the intended family of contrasts.

### 3.2 Continuous analysis

### Center and scale reach

First calculate the average observed reach. Then subtract that value from each observation and divide by 10.

```{r}
#| label: center-and-scale-reach

reach_center_cm <- mean(
  ergo_simple$Reach_cm,
  na.rm = TRUE
)

ergo_simple$Reach_c10 <- (
  ergo_simple$Reach_cm - reach_center_cm
) / 10

print(reach_center_cm)
```

Reach distance was grand-mean centered and expressed in 10-cm units. Consequently, `Reach_c10 = 0` represents the average observed reach, and the main effect of sex compares males and females at that reach distance. Without centering, the sex effect would instead be evaluated at a reach distance of 0 cm, which falls outside the observed range and lacks practical relevance.

### Fit the continuous model

```{r}
#| label: fit-continuous-lmm

model_cont <- lmer(
  Shoulder_Moment_Nm ~ Reach_c10 * Sex +
    (1 | Participant),
  data = ergo_simple,
  REML = TRUE
)

anova_centered <- anova(
  model_cont,
  type = 3,
  ddf = "Satterthwaite"
)

print(anova_centered)
```

The continuous LMM identified significant effects of reach distance and sex, as well as a significant Reach × Sex interaction. The significant interaction indicates that the association between reach distance and shoulder moment differs between males and females. Sex-specific reach slopes were therefore estimated to characterize this relationship within each group.

As in the factor-based example, the complete coefficient table is available from `summary(model_cont)` and is retained in the complete R script.

### Estimate and compare the reach slopes

`emtrends()` estimates the reach slope separately for males and females. Because reach was divided by 10, each slope is the estimated change in shoulder moment for a 10-cm increase in reach.

```{r}
#| label: estimate-sex-specific-reach-trends

reach_trends <- emtrends(
  model_cont,
  ~ Sex,
  var = "Reach_c10",
  lmer.df = "satterthwaite"
)

print(reach_trends)
```

The first follow-up tests each slope against zero. The second directly tests whether the male and female slopes differ. A significant slope for one sex and a nonsignificant slope for the other would not, by itself, establish a difference between the slopes.

```{r}
#| label: test-and-compare-reach-trends

reach_trends_zero <- test(
  reach_trends,
  null = 0,
  adjust = "holm"
)

print(reach_trends_zero)

reach_trends_pair <- pairs(
  reach_trends,
  adjust = "none"
)

print(reach_trends_pair)
```

### Plot the model-estimated relationships

The sex-specific relationships can be visualized by plotting population-level predictions from the fitted LMM over the raw observations. Since reach is represented by a single linear term, the slopes of the prediction lines correspond to the EMTs reported above. The prediction lines should be obtained from the fitted mixed-effects model rather than by adding separate regression lines with `geom_smooth()`, which would fit new models to the raw observations.


```{r}
#| label: plot-continuous-model
#| fig-cap: "Model-estimated relationships between reach distance and shoulder moment by sex. Lines represent population-level predictions from the linear mixed-effects model, shaded bands represent 95% confidence intervals, and points represent raw observations."
#| fig-width: 9
#| fig-height: 5.5
#| out-width: 100%
#| fig-align: center
#| code-fold: true
#| code-summary: "Show continuous-model plotting code"

continuous_model_figure <- sjPlot::plot_model(
  model_cont,
  type = "pred",
  terms = c("Reach_c10 [all]", "Sex"),
  pred.type = "fe",
  show.data = TRUE,
  ci.lvl = 0.95,
  colors = c("#440154", "#7AD151"),
  title = NULL,
  legend.title = "Sex"
) +
  labs(
    x = "Reach Distance (cm)",
    y = "Shoulder Moment (Nm)",
    title = ""
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(continuous_model_figure)
```

The raw observations provide context, while the lines display the population-level relationships estimated by the LMM. The corresponding slopes, confidence intervals, and statistical comparisons should be reported using the EMT results. For nonlinear models, the estimated trend may vary across the predictor range.

## 4. Example 2: Comparing rANOVA and LMMs with missing outcome data

### Introduce two missing outcomes

Copying the data first preserves the complete dataset. The next two commands replace two shoulder-moment values with `NA`, R's marker for a missing value.

```{r}
#| label: introduce-missing-outcomes

ergo_missing <- ergo_simple

ergo_missing$Shoulder_Moment_Nm[
  ergo_missing$Participant == "P03" &
    ergo_missing$Reach_bin == "Far"
] <- NA

ergo_missing$Shoulder_Moment_Nm[
  ergo_missing$Participant == "P04" &
    ergo_missing$Reach_bin == "Near"
] <- NA

print(ergo_missing[is.na(ergo_missing$Shoulder_Moment_Nm), ])
```

The next command attempts rANOVA before the incomplete participants are removed explicitly. `try()` prevents a software error from stopping the tutorial. Depending on the `ez` version, the function may return an error or remove incomplete participants automatically. In either case, the partially observed records are not analyzed as incomplete repeated measures.

```{r}
#| label: attempt-ranova-with-incomplete-data

anova_bins_missing_attempt <- try(
  ezANOVA(
    data = ergo_missing,
    dv = Shoulder_Moment_Nm,
    wid = Participant,
    within = Reach_bin,
    between = Sex,
    detailed = TRUE,
    type = 3
  ),
  silent = TRUE
)

print(anova_bins_missing_attempt)
```

### Prepare complete cases for rANOVA

Conventional rANOVA requires a complete outcome profile for every included participant. The following code identifies the participants with missing outcomes and removes all of their observations.

```{r}
#| label: create-complete-case-data

participants_with_missing <- unique(
  ergo_missing$Participant[
    is.na(ergo_missing$Shoulder_Moment_Nm)
  ]
)

ergo_anova_missing <- ergo_missing[
  !(ergo_missing$Participant %in% participants_with_missing),
]

ergo_anova_missing <- droplevels(ergo_anova_missing)
```

The rANOVA can now be fitted to the reduced complete-case dataset.

```{r}
#| label: fit-complete-case-ranova

anova_bins_missing <- ezANOVA(
  data = ergo_anova_missing,
  dv = Shoulder_Moment_Nm,
  wid = Participant,
  within = Reach_bin,
  between = Sex,
  detailed = TRUE,
  type = 3
)

print(anova_bins_missing$ANOVA)

participants_retained <- unique(
  ergo_anova_missing[c("Participant", "Sex")]
)

print(table(participants_retained$Sex))
```

Removing P03 and P04 leaves 12 complete participants and 36 observations for rANOVA. The complete-case results therefore answer the research questions using less information than was originally collected.

### Fit the LMM using all available outcomes

The LMM is fitted directly to `ergo_missing`. The two missing outcomes are excluded, but the other valid observations from P03 and P04 are retained.

```{r}
#| label: fit-factor-lmm-with-missing-data

model_bins_missing <- lmer(
  Shoulder_Moment_Nm ~ Reach_bin * Sex +
    (1 | Participant),
  data = ergo_missing,
  REML = TRUE,
  na.action = na.exclude
)

anova_bins_lmm_missing <- anova(
  model_bins_missing,
  type = 3,
  ddf = "Satterthwaite"
)

print(anova_bins_lmm_missing)
```

The factor-based LMM retains 40 observed outcomes from all 14 participants. Reach condition, sex, and their interaction remain statistically significant, although the estimates and test statistics differ slightly from the complete-data analysis because two outcomes are now unavailable.

As in the complete-data analysis, EMMs describe the model-estimated shoulder moment for each reach condition within each sex.

```{r}
#| label: estimate-factor-emms-with-missing-data

emm_bins_missing <- emmeans(
  model_bins_missing,
  ~ Reach_bin | Sex,
  lmer.df = "satterthwaite"
)

print(emm_bins_missing)
```

The estimated means continue to increase from Near to Far for both sexes. Pairwise comparisons are then used to determine which reach conditions differ within each sex. Tukey adjustment is applied to the three comparisons in each family.

```{r}
#| label: compare-factor-emms-with-missing-data

pairs_bins_missing <- pairs(
  emm_bins_missing,
  adjust = "tukey"
)

print(pairs_bins_missing)
```

All three reach conditions differ significantly among males. Among females, Near differs from both Mid and Far, whereas Mid and Far do not differ significantly. These results are specific to the observed incomplete dataset and differ slightly from the complete-data comparisons.

The same approach could also be used for the continuous reach model.

### Compare how much data each analysis retains

The practical difference between the approaches can be summarized by comparing the number of participants and observed outcomes included in each analysis.

```{r}
#| label: compare-data-retention

data_retained <- data.frame(
  Analysis = c(
    "Complete-case rANOVA",
    "Factor-based LMM"
  ),
  Participants_Retained = c(
    length(unique(ergo_anova_missing$Participant)),
    length(unique(
      ergo_missing$Participant[
        !is.na(ergo_missing$Shoulder_Moment_Nm)
      ]
    ))
  ),
  Observations_Retained = c(
    nrow(ergo_anova_missing),
    nobs(model_bins_missing)
  )
)

print(data_retained)
```

The complete-case rANOVA retained 12 participants and 36 observations, whereas the factor-based LMM retained all 14 participants and 40 observed outcomes. This comparison illustrates information retention, not a guarantee that LMMs will produce significant results or eliminate bias. Valid inference still depends on appropriate model specification and defensible assumptions about the missing-data mechanism.

## 5. Model diagnostics

Model diagnostics should be examined for every LMM used for inference. The `check_model()` function produces several figures for examining residual patterns, normality, constant variance, and potentially influential observations, as discussed in the article. The commands are not executed automatically during rendering because each call produces a large set of diagnostic figures. Readers should run each command separately in an interactive R session and review the results before interpreting or reporting the corresponding model.

```{r}
#| label: model-diagnostics
#| eval: false
#| code-fold: true
#| code-summary: "Show model diagnostic code"

check_model(model_bins)
check_model(model_cont)
check_model(model_bins_missing)
```

## 6. Scripts included with the tutorial

`R/00_Complete_Analysis.R` contains the complete workflow in one file. The remaining scripts contain one example each and repeat their setup so they can be run independently. A beginner can therefore start with the annotated tutorial and later use the shorter scripts as analysis templates.
